---
title: "Import example2"
error: true
embed-resources: true
---


```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(sf)       # for some map stuff, Julian will talk more about the details
d <- read_excel("data-raw/Coordinates for the map.xlsx")
# You could also try this, if you do not have the excel file on your local computer:
# d <- read_csv("https://heima.hafro.is/~einarhj/data/some_messy_map_coordinates.csv")
```


Let's take a peek of the data:
```{r}
d |> glimpse()
```


We take note that the coordinates are with all this fancy stuff like "degrees", "minutes" and "seconds" notation. Since we often come across such data we may try to generate first a function:

```{r}
# This function removes all characters except numerics (0, 1, 2, ... 9) and the "." and puts
#  a single space instead, then separates out the degrees, minutes and seconds and the
#  calculates the decimal degrees
dms2decimmals <- function(x) {
  tibble::tibble(x = x) |> 
    dplyr::mutate(x = stringr::str_replace_all(x, "[^0-9.-]", " "),
           x = stringr::str_squish(x)) |> 
    tidyr::separate(x, into = c("dd", "mm", "ss"), convert = TRUE, sep = " ") |> 
    dplyr::mutate(lon = dd + mm / 60 + ss / 3600) |> 
    dplyr::pull(lon)
}
```

Now let's use this function in a code flow, first just a demo to see the output:

```{r}
d |> 
  mutate(lat = dms2decimmals(Latitude),
         lon = dms2decimmals(Longitude)) |> 
  knitr::kable()
```

Now, let's look carefully the lat and the lon data. What do you notice is wrong given the orignal data??

The problem here is that the function always gives us positive numbers. But in decimal degrees if the latitude is in the southern hemisphere that should give us a negative decimal degrees. Same would be if the longitude is west of the Greenwich line.

Let's fix the function above to take that into account:

```{r}
# This function removes all characters except numerics (0, 1, 2, ... 9) and the "." and puts
#  a single space instead, then separates out the degrees, minutes and seconds and the
#  calculates the decimal degrees
# This ammended function now takes care of if one has N, S, E, W in the original values
dms2decimmals <- function(x) {
  
  # get only letters
  quarter <- x |> stringr::str_extract("[A-Z]")
  x <- 
    tibble::tibble(x = x) |> 
    dplyr::mutate(x = stringr::str_replace_all(x, "[^0-9.-]", " "),
           x = stringr::str_squish(x)) |> 
    tidyr::separate(x, into = c("dd", "mm", "ss"), convert = TRUE, sep = " ") |> 
    dplyr::mutate(lon = dd + mm / 60 + ss / 3600) |> 
    dplyr::pull(lon)
  case_when(quarter == "S" | quarter == "W" ~ -x,
            .default = x)
}
```

Now let's test again:

```{r}
d |> 
  mutate(lat = dms2decimmals(Latitude),
         lon = dms2decimmals(Longitude)) |> 
  knitr::kable()
```

If we are happy we can overwrite the variables and store the object (read: table) under the same name.

```{r}
d <-
  d |> 
  mutate(Latitude = dms2decimmals(Latitude),
         Longitude = dms2decimmals(Longitude))
d
```

Let's clean the names (we could also have done this right after importing the file):

```{r}
d <- 
  d |> 
  clean_names()
```

Now we could do a plot:

```{r}
d |> 
  ggplot(aes(longitude, latitude)) +
  geom_point(colour = "red") +
  geom_text(aes(label = name_of_island), 
            angle = 45) +
  coord_quickmap()
```

Almost there, but we may want a background map of Lake Victoria. Let's get one:

```{r}
# https://www.nature.com/articles/s41597-022-01742-3
# https://dataverse.harvard.edu/dataverse/GLWNB-2020
lv <- 
  read_sf("https://heima.hafro.is/~einarhj/data/lake_victoria_shoreline.gpkg")
```

And redo the plot:

```{r}
ggplot() +
  geom_sf(data = lv,
          fill = "lightblue") +
  geom_point(data = d,
             aes(longitude, latitude),
             colour = "red") +
  #geom_text(data = d,
  #          aes(longitude, latitude, label = name_of_island), 
  #          angle = 0) +
  coord_sf(xlim = c(32, 33), ylim = c(-0.2, 0.4)) +
  labs(x = NULL, y = NULL)
```

